\hydrogen{} and \carbon{} Network Strategies for the European Energy System

### 2024-08-31 report/submissions/journal/paper/paper.tex
add appendix section about higher sequestration potentials

In the literature, the two network approaches and underlying technologies have been discussed in several publications, all of which, however, dealt with the isolated aspects~\cite{bakkenLinearModelsOptimization2008,morbeeOptimisedDeploymentEuropean2012,stewartFeasibilityEuropeanwideIntegrated2014,oeiModelingCarbonCapture2014,elahiMultiperiodLeastCost2014,burandtDecarbonizingChinaEnergy2019,middletonSimCCSOpensourceTool2020,bjerketvedtOptimalDesignCost2020,weiProposedGlobalLayout2021,damoreOptimalDesignEuropean2021,becattiniCarbonDioxideCapture2022,neumannBenefitsHydrogenNetwork2022}. Such techno-economic models, in contrast for example to Integrated Assessment Models, can account for the spatial distribution of carbon sources and sinks, which are crucial to provide a holistic view of the energy system and its technological interactions. Neumann et al.~\cite{neumannBenefitsHydrogenNetwork2022} demonstrated the interchangeability of hydrogen and electricity grid expansion in a climate-neutral European energy system at the expense of higher investments. The underlying, highly-resolved model encompasses the most relevant sectors, however it neglects the option to transport carbon across regions.
Morbee et al.~\cite{morbeeOptimisedDeploymentEuropean2012} optimizes the topology and capacity of a \carbon{} network in Europe, but only considers the power sector without co-optimizing renewable deployment. This limited sectoral scope cannot capture important dynamics of carbon management, since it neglects sectors like industry that will need to handle most \carbon{} in the future.
The European Commission's Joint Research Center analysed the possible topology of a \carbon{} network in 2024 \cite{jrc2024}, but didn't include the interaction of  \carbon{} sources and sinks with the rest of the energy system.
A comprehensive example is found in~\cite{becattiniCarbonDioxideCapture2022}, which presents a mixed-integer model to optimize the time-evolution of a \carbon{} transport system in Switzerland, connected to a remote sequestration site in Norway. However, this limited spatial scope fails to consider other sequestration sites and the co-benefits from connecting the \carbon{} network to neighboring countries.
Other high-resolution energy system models do not feature detailed CU technologies and \carbon{} transport. Euro-Calliope, for example, does not feature carbon nor hydrogen transport, and thus forces generic synthetic fuel production to rely on captive electrolysis and DAC in the same region~\cite{pickeringDiversityOptionsEliminate2022}.

@@

To investigate the competing transport dynamics between \carbon{} and \hydrogen{} networks, we contrast the following four scenarios:
%
\begin{itemize}
    \item[] \textit{Baseline:} Neither \carbon{} nor \hydrogen{} networks are constructed.
    \item[] \textit{\carbongrid{}:} Only the \carbon{} network is developed.
    \item[] \textit{\hydrogengrid{}:} Only the \hydrogen{} network is developed.
    \item[] \textit{Hybrid:} Both \carbon{} and \hydrogen{} networks are developed.
\end{itemize}

@@

\subsection*{Higher sequestration limits reintroduce fossil fuels}
\label{sec:higher_sequestration}
Carbon management and hard-to-abate sectors of the optimized system are strongly driven by the model's limited sequestration potential of 200~Mt/a. In the Hybrid scenario, this limitations imposes a shadow-price of 225€/t to the system, meaning the system would gain 225€ from an additional tonne of sequestration potential. As soon as the sequestration limit is increased, the model starts to reintroduce fossil fuels as a primary energy source in the the aviation and industrial sectors.

@@

In the \carbon{} and \hydrogen{} scenarios, this slight shift in CU deployment results in a lower buildout of hydrogen pipelines, especially from the Iberian Peninsula to Central Europe (0.5~bn€/a). Nonetheless, 80\% of the hydrogen topology which is built under net-zero emissions is also built under the assumption of net-negative emissions. Simultaneously, under net-negative emissions the \carbon{} network also includes routes from Central and Eastern Europe to the shores (1.2~bn€/a). The benefits of \carbon{} transport and sequestration with tighter emission targets are also evident when comparing the other scenarios. The carbon network under the net-negative \carbonscenario{} is larger than under the net-zero \carbonscenario{} and transports more carbon from point sources to sequestration sites rather than to CU facilities. In the \hydrogenscenario{}, where all carbon for CU and CS must come from DAC, the hydrogen transport topology remains almost identical to the net-zero scenarios (for detailed cost contributions, refer to sections~\ref{sec:cost_comparison} and~\ref{sec:operation_nn} in the Appendix).

@@

Limitations of the study, such as the neglect of synthetic fuel imports and assumptions on biomass potentials, are discussed in the Appendix in Section~\ref{sec:limitations}. A sensitivity analysis of the costs of \carbon{} pipelines, which show negligible effects on the results, can be found in the Appendix in Section~\ref{sec:subsidy}.

@@

>>>>

In the literature, the two network approaches and underlying technologies have been discussed in several publications, all of which, however, dealt with the isolated aspects~\cite{bakkenLinearModelsOptimization2008,morbeeOptimisedDeploymentEuropean2012,stewartFeasibilityEuropeanwideIntegrated2014,oeiModelingCarbonCapture2014,elahiMultiperiodLeastCost2014,burandtDecarbonizingChinaEnergy2019,middletonSimCCSOpensourceTool2020,bjerketvedtOptimalDesignCost2020,weiProposedGlobalLayout2021,damoreOptimalDesignEuropean2021,becattiniCarbonDioxideCapture2022,neumannBenefitsHydrogenNetwork2022}. Such techno-economic models, in contrast for example to Integrated Assessment Models, can account for the spatial distribution of carbon sources and sinks, which are crucial to provide a holistic view of the energy system and its technological interactions. Neumann et al.~\cite{neumannBenefitsHydrogenNetwork2022} demonstrated the interchangeability of hydrogen and electricity grid expansion in a climate-neutral European energy system at the expense of higher investments. The underlying, highly-resolved model encompasses the most relevant sectors, however it neglects the option to transport carbon across regions.
Morbee et al.~\cite{morbeeOptimisedDeploymentEuropean2012} optimizes the topology and capacity of a \carbon{} network in Europe, but only considers the power sector without co-optimizing renewable deployment. This limited sectoral scope cannot capture important dynamics of carbon management, since it neglects sectors like industry that will need to handle most \carbon{} in the future.
The European Commission's Joint Research Center analysed the possible topology of a \carbon{} network in 2024~\cite{jrc2024}, but didn't include the interaction of  \carbon{} sources and sinks with the rest of the energy system.
A comprehensive example is found in~\cite{becattiniCarbonDioxideCapture2022}, which presents a mixed-integer model to optimize the time-evolution of a \carbon{} transport system in Switzerland, connected to a remote sequestration site in Norway. However, this limited spatial scope fails to consider other sequestration sites and the co-benefits from connecting the \carbon{} network to neighboring countries.
Other high-resolution energy system models do not feature detailed CU technologies and \carbon{} transport. Euro-Calliope, for example, does not feature carbon nor hydrogen transport, and thus forces generic synthetic fuel production to rely on captive electrolysis and DAC in the same region~\cite{pickeringDiversityOptionsEliminate2022}.

@@

To investigate the competing transport dynamics between \carbon{} and \hydrogen{} networks, we contrast the following four scenarios:
\begin{table}[]
    \resizebox{\textwidth}{!}{%
        \begin{tabular}{@{}lcc@{}}
            \toprule
            \textbf{Scenario} & H$_2$ Network & CO$_2$ Network \\ \midrule
            \textbf{Baseline} & \xmark          & \xmark           \\
            \textbf{CO$_2$-Grid} & \xmark          & \checkmark            \\
            \textbf{H$_2$-Grid}  & \checkmark          & \xmark           \\
            \textbf{Hybrid}   & \checkmark           & \checkmark            \\ \bottomrule
        \end{tabular}%
    }
\caption{Overview of the four different scenarios.}
\label{tab:overview_scenarios}
\end{table}

@@

@@

In the \carbon{} and \hydrogen{} scenarios, this slight shift in CU deployment results in a lower buildout of hydrogen pipelines, especially from the Iberian Peninsula to Central Europe (0.5~bn€/a). Nonetheless, 80\% of the hydrogen topology which is built under net-zero emissions is also built under the assumption of net-negative emissions. Under net-negative emissions the \carbon{} network also includes routes from Central and Eastern Europe to the shores (1.2~bn€/a). The benefits of \carbon{} transport and sequestration with tighter emission targets are also evident when comparing the other scenarios. The carbon network under the net-negative \carbonscenario{} is larger than under the net-zero \carbonscenario{} and transports more carbon from point sources to sequestration sites rather than to CU facilities. In the \hydrogenscenario{}, where all carbon for CU and CS must come from DAC, the hydrogen transport topology remains almost identical to the net-zero scenarios (for detailed cost contributions, refer to sections~\ref{sec:cost_comparison} and~\ref{sec:operation_nn} in the Appendix).

@@

Limitations of the study, such as the neglect of synthetic fuel imports and assumptions on biomass potentials, are discussed in the Appendix in Section~\ref{sec:limitations}. A sensitivity analysis of the costs of \carbon{} pipelines and costs of electrolyzers, which show relatively small effects on the results, can be found in the Appendix in Section~\ref{sec:subsidy}. As the cost optimum is affected by the sequestration potential limits, a more detailed analysis of a relaxed sequestration limit is presented in the Appendix in Section~\ref{sec:higher_sequestration}.

@@

\subsection*{Higher sequestration limits}
\label{sec:higher_sequestration}
Carbon management in the optimised energy system, particularly in the hard to abate sectors, is heavily influenced by the model's sequestration potential, which is capped at 200~Mt/year. Lifting the sequestration limit leads to cost benefits and a shift in primary energy supply (see Fig.~\ref{fig:sequestration_cost_bar}). Fossil oil and gas increasingly provide critical energy services in the aviation, industrial and heating sectors. Their emissions are either directly capture at the point source and transported to be sequestered or offset large DAC facilities close to coastal sequestration sites. At the same time, the role of the \hydrogen{} network is reduced from 57~TWkm to 20~TWkm due to the less CU required.
At 800~Mt/a this already reduces wind and solar investments by a third and total costs by 9.1\%, roughly aligning with the findings in~\cite{hofmannDesigningCO2Network2023}.

As the sequestration capacity increases, the cost savings from additional sequestration begin to plateau. At 200Mt/a, the shadow price of the sequestration constraint shows that each additional tonne saves 225€/t, but at 800Mt/a, this saving drops to 51€/t. Extrapolating this behaviour, saturation of sequestration can be expected at around 1200~Mt/a.
